Synthesizing Symmetric Lenses

### 2018-07-09 papers/SymmetricLenses/main.tex
related

% begin related-work
\section{Relationship To Classical Symmetric Lenses}
\label{sec:relationship}
\paragraph*{Formalization} We have reformulated the problem addressed by classical
symmetric lenses with simple symmetric lenses. A classical symmetric lens
$\Lens$ between $X$ and $Y$ consists of 4 components: a complement $C$, a
designated element $init \in C$, and two functions for propagating changes in
each format to the other:
\begin{enumerate}
\item $\PutRSym \OfType X \times C \rightarrow Y \times C$
\item $\PutLSym \OfType Y \times C \rightarrow X \times C$
\end{enumerate}
In this formulation, the components of each format that aren't shared with the
other format are stored in the complement. When one format is edited, the \PutR
or \PutL function stitches together that edited data with the data for the other
side, stored in the complement.  The designated element, $init$, is used as the
initial value of $C$, and specifies default behavior when there isn't existing
data to stitch with.  In the example in Figure~\ref{fig:minimized-representations} the complement would consist of a list of pairs with each pair consisting of an integral salaray value and the name of a company.

These components only make sense for synchronization when they satisfy
round-tripping laws.  This guarantees that if an identity edit is propagated, no
data gets changed.
\begin{equation}
  \tag{\PutRL}
  \frac{\PutRSymOf{(x,c)} = (y,c'')}{\PutLSymOf{(y,c'''')} = (x,c')}
\end{equation}
\begin{equation}
  \tag{\PutLR}
  \frac{\PutLSymOf{(y,c)} = (x,c')}{\PutRSymOf{(x,c')} = (y,c')}
\end{equation}

Furthermore, two classical symmetric lenses are equivalent if they always output the same
formats on any sequence of edits. Formally, given a lens $\Lens \in X
\leftrightarrow Y$, an \emph{edit} for $\Lens$ is a member of $X + Y$. Consider
the function $apply$, which, given a lens and an element of that lens's
complement, is a function from sequences of edits to sequences of edits. If
$apply(\Lens,c,es) = es'$, then that means, starting with complement $c$, after
edit $es_i$, the lens $\Lens$ synchronizes the other data to $es'_i$.
\[
  \inferrule
  {
  }
  {
    apply(\Lens,c,[]) = []
  }
\]
\[
  \inferrule
  {
    \Lens.putr(x,c) = (y,c')\\
    apply(\Lens,c',es) = es'
  }
  {
    apply(\Lens,c,(\InLOf{x})::es) = (\InROf{y})::es'
  }
\]
\[
  \inferrule
  {
    \Lens.putl(y,c) = (x,c')\\
    apply(\Lens,c',es) = es'
  }
  {
    apply(\Lens,c,(\InROf{y})::es) = (\InLOf{x})::es'
  }
\]

@@

\section{Related Work}
\label{sec:related}
There is related work in a number of directions. There is related work in string
transformation synthesis and type-directed synthesis. We also refer the reader
to a recent, general survey on synthesis~\ref{?}.

\subsection{String Transformation Synthesis}
Much of the research on string transformation synthesis is on synthesizing
string transformations from input/output
specifications~\cite{flashfill,flashextract,?}. While this provides an easier
interaction model with the user than this work, it requires the algorithm to
learn the data formats from the examples. This makes the task much more
difficult for the synthesis algorithm, particularly in addressing iterations and
disjunctions. Prior work has shown that learning from examples and data format
descriptions enables synthesis of much more complex transformations than are
permitted by examples alone. Recent work has gone into making these algorithms
more robust by delaying the selection of program until seeing what further
inputs the program is run on~\cite{?}. By doing this, the synthesis algorithm
can learn additional information about the formats from the use of the program.

In many of these synthesis algorithms the correct transformation is chosen by
principals like ``minimize constants''~\cite{?,?,?} and ``most specific
generalization''~\cite{?,?,?}. We are the first work to our knowledge to use an
information-theoretic measure for finding the correct transformation. However,
much of our metrics are compatible with existing metrics: avoiding \Disconnect
lenses minimizes constants and creates more specfic transformations (\Disconnect
lenses permit more synchronized data).

Optician~\ref{?} synchronizes string transformation functions, but prior
versions were very restrictive in which transformations were permitted.
Originally, Optician only synchronized transformations between formats in
bijective correspondence, but few formats satisfy that constraint. Recent
enhancements have permitted synthesis of more expressive lenses, quotient
bijective lenses. These enhancements were made by strengthening the
specifications from regular expressions to QREs, regular expressions annotated
with information on normal forms of the language. This enhancement permits
synthesizing transformations between formats whose normal forms are in bijective
correspondence. While we have not integrated our system to permit QRE
specifications, we see no reason the two systems cannot be integrated.

Other work synthesizing invertible transformations was done by inverting a
previously defined transformation.  This has been done by exploiting
parametricity to invert transformations~\cite{?}, and by inverting symbolic
finite automata~\cite{?}.

\subsection{Type-Directed Synthesis}

Type-directed synthesis uses type information to tighten specifications and
reduce the number of possible programs. Typically, type-directed synthesis
techniques use the types to orient the search space.  Our work interestingly
deals with types with many semantic equivalences on them. InSynth~\cite{?}
deals with the same problem when trying to synthesize transformations involving
library functions. InSynth addresses this in a similar manner to this work: by
normalizing the types we don't need to search through the equivalences on those
types.

>>>>

% begin related-work
\section{Related Work}

In this paper, we have designed and implemented a new, pragmatic
formulation of symmetric lenses as well as new program synthesis
techniques.  In this section, we consider the relationship with
classical symmetric lenses in depth and also comment on related
program synthesis techniques.

\subsection{Relationship with Classical Symmetric Lenses}
\label{sec:relationship}
%\paragraph*{Formalization}
%We have reformulated the problem addressed by classical
%symmetric lenses with simple symmetric lenses.
A classical symmetric lens
$\Lens$ between $X$ and $Y$ consists of 4 components: a complement $C$, a
designated element $init \in C$, and two functions,
$\PutRSym$ and $\PutLSym$, that propagate changes in one format to the other.
\begin{enumerate}
\item $\PutRSym \OfType X \times C \rightarrow Y \times C$
\item $\PutLSym \OfType Y \times C \rightarrow X \times C$
\end{enumerate}
In this formulation, data unique to each side are stored in the complement.
When one format is edited, the \PutR
or \PutL function stitches together the edited data with data
stored in the complement.  The  $init$ element is used as the
initial value of $C$ and specifies default behavior when data is missing.
For instance, to implement the
scenario in Figure~\ref{fig:minimized-representations}, the complement
would consist of a list of pairs of salary and company name.

Classical symmetric lenses satisfy the following equational laws.
\begin{equation}
  \tag{\PutRL}
  \frac{\PutRSymOf{(x,c)} = (y,c'')}{\PutLSymOf{(y,c'''')} = (x,c')}
\end{equation}
\begin{equation}
  \tag{\PutLR}
  \frac{\PutLSymOf{(y,c)} = (x,c')}{\PutRSymOf{(x,c')} = (y,c')}
\end{equation}

Two classical symmetric lenses are equivalent if they output the same
formats given any sequence of edits. Formally, given a lens
$\Lens \in X \leftrightarrow Y$,
an \emph{edit} for $\Lens$ is a member of $X + Y$. Consider
the function $apply$, which, given a lens and an element of that lens's
complement, is a function from sequences of edits to sequences of edits. If
$apply(\Lens,c,es) = es'$, then that means, starting with complement $c$, after
edit $es_i$, the lens $\Lens$ synchronizes the other data to $es'_i$.
\[
  \inferrule
  {
  }
  {
    apply(\Lens,c,[]) = []
  }
\]
\[
  \inferrule
  {
    \Lens.putr(x,c) = (y,c')\\
    apply(\Lens,c',es) = es'
  }
  {
    apply(\Lens,c,(\InLOf{x})::es) = (\InROf{y})::es'
  }
\]
\[
  \inferrule
  {
    \Lens.putl(y,c) = (x,c')\\
    apply(\Lens,c',es) = es'
  }
  {
    apply(\Lens,c,(\InROf{y})::es) = (\InLOf{x})::es'
  }
\]

@@

\subsection{Data Transformation Synthesis}
Over the past decade, the programming languages community has explored
the synthesis of programs from a wide variety of angles.  One of
the key ideas is typically to narrow the program search space by focusing
on a specific domain, as we have done here.
%, and as has been advocated
%more generally under the rubric of syntax-guided synthesis~\cite{sygus}.

Automation of string transformations, in particular, has been the focus
of much prior attention.  For example, Gulwani and others work on FlashFill
generates one-way spreadsheet transformations from input/output
examples~\cite{flashfill,flashextract}.  On the one hand,
FlashFill is easier to use because one need not specify the type of
the data being transformed.  On the other hand, this type information
makes it possible to transform more complex formats---Flash Fill
does not synthesize programs with nested loops, for instance,
and hence is incapable of synthesizing the majority of our
benchmarks, even in one direction~\cite{flashfill}.
%% Flashfill, designed
%% for transformation of the cells of spreadsheets, does not scale to
%% more complex formats
%% While this provides an easier
%% interaction model with the user than this work, it requires the algorithm to
%% learn the data formats from the examples. This makes the task much more
%% difficult for the synthesis algorithm, particularly in addressing iterations and
%% disjunctions. Prior work has shown that learning from examples and data format
%% descriptions enables synthesis of much more complex transformations than are
%% permitted by examples alone. Recent work has gone into making these algorithms
%% more robust by delaying the selection of program until seeing what further
%% inputs the program is run on~\cite{?}. By doing this, the synthesis algorithm
%% can learn additional information about the formats from the use of the program.

All pragmatic synthesis algorithms use heuristics of one kind or another.  One of the
goals of the current paper is to try to ground those heuristics in a broader theory,
information theory,
with the hope that this theory may inform future design decisions and help us understand
heuristics crafted in other tools and possibly in other domains.  For instance, we
speculate that some of the heuristics used in
FlashFill (to take one well-documented example) may be connected to some of the principles
laid out here.  For instance, FlashFill prioritizes the substring constructor over the
constant string constructor when ranking possible programs. Such a choice may be understood
in information-theoretic terms as the translation from target back to source would require
more information.  Said another way, our information theoretic metrics measure injectivity
(in both directions, in our case, to measure bijectivity) but applied in one direction, as
is the case for FlashFill, would avoid the highly injective constant functions.  Beyond
constants, another FlashFill heuristic prefers ``wider'' character classes (or token classes)
over ``narrower'' ones.  Selecting a wider set of strings to copy from left-to-right, viewed
in the reverse direction, from right-to-left, results in a function that is closer to being
injective, as measured by information theory.  These are just a few of the heuristics used
in Flashfill; more research would be required (and public disclosure of the details necessary)
to understand the connections fully.

%% In many of these synthesis algorithms the correct transformation is chosen by
%% principals like ``minimize constants''~\cite{?,?,?} and ``most specific
%% generalization''~\cite{?,?,?}. We are the first work to our knowledge to use an
%% information-theoretic measure for finding the correct transformation. However,
%% much of our metrics are compatible with existing metrics: avoiding \Disconnect
%% lenses minimizes constants and creates more specfic transformations (\Disconnect
%% lenses permit more synchronized data).

As discussed in the introduction,
the Optician tool~\cite{optician,maina+:quotient-synthesis} was a building block for
our work.  Optician synthesized bijective transformations~\cite{optician} and
bijections modulo quotients~\cite{maina+:quotient-synthesis}, but could
not synthesize more complex bidirectional transformations where one format contains
important information not present in the other---a common situation in the real
world.  From a technical perspective, the first key novelty in our work involves the definition, theory,
analysis, and implementation of a new class of simple symmetric lenses, designed
for synthesis.  The second key technical innovation involves the use of stochastic
regular expressions and information theory to guide the search for program transformations.
As mentioned earlier, we believe such information theoretic techniques may have broad utility
in helping us understand how to formulate a search for a data transformation function.
%% The central contributions
%% only synchronized transformations between formats in
%% bijective correspondence, but few formats satisfy that constraint. Recent
%% enhancements have permitted synthesis of more expressive lenses, quotient
%% bijective lenses. These enhancements were made by strengthening the
%% specifications from regular expressions to QREs, regular expressions annotated
%% with information on normal forms of the language. This enhancement permits
%% synthesizing transformations between formats whose normal forms are in bijective
%% correspondence. While we have not integrated our system to permit QRE
%% specifications, we see no reason the two systems cannot be integrated.

While our tool uses types and examples to specify invertable transformations, other
tools have been shown to synthesize a backwards transformation from ordinary code designed
to implement the forwards transformation.  For instance, Hu and
D'Antoni~\cite{program-inversion-symbolic-transducers} show how to invert transformations
using symbolic finite automata, and
Voigtl\"{a}nder~\cite{bidirectionalization-for-free} demonstrates how to
construct reverse transformations from forwards transformations
by exploiting parametricity properties.  These kinds of tools are very useful, but in different
circumstances from ours (namely, when one already has the code to implement one direction of
the transformation).

%% Other work synthesizing invertible transformations was done by inverting a
%% previously defined transformation.  This has been done by exploiting
%% parametricity to invert transformations~\cite{}, and by inverting symbolic
%% finite automata~\cite{?}.

%% \subsection{Type-Directed Synthesis}

%% Type-directed synthesis uses type information to tighten specifications and
%% reduce the number of possible programs. Typically, type-directed synthesis
%% techniques use the types to orient the search space.  Our work interestingly
%% deals with types with many semantic equivalences on them. InSynth~\cite{?}
%% deals with the same problem when trying to synthesize transformations involving
%% library functions. InSynth addresses this in a similar manner to this work: by
%% normalizing the types we don't need to search through the equivalences on those
%% types.
